---
title: 'Data wrangling - Extract , Transform and Load Data'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Imports:
1. library(dplyr)          # data wrangling
2. library(ggplot2)        # visualization
3. library(tidyverse)      # tidyverse
4. library(tibble)         # tibble
5. library(tsibble)        # tsibble
6. library(knitr)          # knitr to display table

```{r, include=FALSE}
library(dplyr)          # data wrangling
library(ggplot2)        # visualization
library(tidyverse)      # tidyverse
library(tibble)         # tibble
library(tsibble)        # tsibble
library(knitr)          # knitr to display table

```

## 1. Reading Dataset into R:
```{r}

#set working directory to pull data

setwd("C:/Users/Titan/Desktop/UVA/UVA_AA_Semesters/UVA_Spring2022/SYS5581_TimeSeries/TimeSeries_Datasets/Mullick-Tahsin/MHEALTHDATASET/")


# Read data in as a tidy dataframe
df <- read.table("subject10.txt", sep="")

```
## 2. Organize the into Tidy data
```{r}

# setting column names for the data file based on the datasource

colnames(df) <- c('acc_chest_X','acc_chest_Y','acc_chest_Z','ecg_L1','ecg_l2','acc_leftankle_X','acc_leftankle_Y','acc_leftankle_Z','gyro_leftankle_X','gyro_leftankle_Y','gyro_leftankle_Z','mag_leftankle_X','mag_leftankle_Y','mag_leftankle_Z', 'acc_rlarm_X','acc_rlarm_Y','acc_rlarm_Z','gyro_rlarm_X','gyro_rlarm_Y','gyro_rlarm_Z','mag_rlarm_X','mag_rlarm_Y','mag_rlarm_Z','activity')

```


### Looking at the head of the dataset to ensure dataset structure as per Tidy data
```{r}

#View(head(df,n=10))

knitr::kable(df[1:5,],"pipe" , caption = "Subject 10 Activity Sensor Data")

```
## 3. Clean the data and perform various data quality checks:

### Inspcting  type of variables in the dataframe
```{r}
typeof(df)
str(df)
```
### Converting the 'activity' column into type-factor:
```{r}
df[,'activity']<-factor(df[,'activity'])
typeof(df$activity)
class(df$activity)
```


## Checking for missing values per column:

```{r}
sapply(df, function(x) sum(is.na(x)))
```

### Adding Data Explorer library
```{r, include = FALSE}
library(DataExplorer)
```




### Missing value plot
```{r}
plot_missing(df)
```
### plot histogram of the variables
```{r}
plot_histogram(df)
```

## 4. Checking if data needs to be converted to tsibble
####  The data I have is does not contain a particular timestamp but rather is a sequence of sensor readings 50Hz apart as such converting it to tsibble is not required



## 5. Generating a graph of the acceleration of the chest along X axis grouped on activity
```{r}
ggplot(df, 
       aes(x=acc_chest_X, group = activity, fill= activity)) +
  geom_density(alpha=.5)
```

